Inferring the effective start dates of non-pharmaceutical interventions during \covid outbreaks

### 2020-05-24 ms/ms.tex
UO comments

% NatBib
\usepackage[super]{natbib}
%\usepackage[numbers,super,sort&compress]{natbib}
%\usepackage[round,colon,authoryear]{natbib}
\renewcommand{\bibsection}{}
%\renewcommand{\bibfont}{\small}

@@

During February and March 2020, several countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules to control the \covid pandemic caused by the \sars virus.
Overall, these interventions seem to have successfully reduced the spread of the pandemic.
We hypothesize that the official and effective start date of such interventions can significantly differ, for example due to slow diffusion of guidelines in the population, or due to unpreparedness of the authorities and the public.
We use an SEIR epidemiological model and an MCMC inference framework to estimate the effective start of NPIs in several countries, and compare these effective dates to the official dates.
We report our finding of both late and early effects of NPIs, and discuss potential causes and consequences of our results.

@@

The COVID-19 pandemic has resulted in implementation of extreme non-pharmaceutical interventions (NPIs) in many affected countries. These interventions, from social distancing to lockdowns, are applied in a rapid and widespread fashion.
The NPIs are designed and assessed using epidemiological models, which follow the dynamics of the viral infection to forecast the effect of different mitigation and suppression strategies on the levels of infection, hospitalization, and fatality.
These epidemiological models usually assume that the effect of NPIs on disease transmission begins at the officially declared date (e.g.~\citet{Li2020,Gatto2020,Flaxman2020}).

Adoption of public health recommendations is often critical for effective response to infectious diseases, and has been studied in the context of HIV~\citep{Kaufman2014} and vaccination~\citep{Dunn2015,Wiyeh2018}, for example.
However, behavioral and social change does not occur immediately, but rather requires time to diffuse in the population through media, social networks, and social interactions. 
Moreover, compliance to NPIs may differ between different interventions and between people.
For example, in a survey of 2,108 adults in the UK during Mar 2020, \citet{Atchison2020} found that those over 70 years old were more likely to adopt social distancing than young adults (18-34 years old), and that those with lower income were less likely to be able to work from home and to self-isolate.
Similarly, compliance to NPIs may be impacted by personal experiences. \citet{Smith2020} have surveyed 6,149 UK adults in late April and found that people who believe they have already had \covid are more likely to think they are immune, and less likely to comply with social distancing measures. 
Compliance may also depend on risk perception as perceived by the the number of domestic cases or even by reported cases in other regions and countries.
Interestingly, the perceived risk of \covid infection has likely caused a reduction in the number of influenza-like illness cases in the US starting from mid-February~\citep{Zipfel2020}.

Here, we hypothesize that there is a significant difference between the official start of NPIs and their adoption by the public and therefore their effect on transmission dynamics.
We use a \textit{Susceptible-Exposed-Infected-Recovered} (SEIR) epidemiological model and \textit{Markov Chain Monte Carlo} (MCMC) parameter estimation framework to estimate the effective start date of NPIs from publicly available COVID-19 case data in several geographical regions.
We compare these estimates to the official dates and find both late and early effects of NPIs on \covid transmission dynamics.
We conclude by demonstrating how differences between the official and effective start of NPIs can confuse assessments of the effectiveness of the NPIs in a simple epidemic control framework. % TODO explain why important

@@

%%% Data %%%%
\paragraph*{Data.} 
We use daily confirmed case data $\vec{X}=(X_1, \ldots, X_T)$ from several different countries. These incidence data summarize the number of individuals $X_t$ tested positive for SARS\=/CoV\=/2 RNA (using RT-qPCR) at each day $t$.
Data for Wuhan, China retrieved from \citet{Pei2020}, data for 11 European countries retrieved from \citet{Flaxman2020}. 
Regions in which there were multiple sequences of days with zero confirmed cases (e.g. France), we cropped the data to begin with the last sequence so that our analysis focuses on the first sustained outbreak rather than isolated imported cases. 
For dates of official NPI dates see \autoref{table:NPI_dates}.

@@

Susceptible ($S$) individuals become exposed due to contact with reported or unreported infected individuals ($I_r$ or $I_u$) at a rate $\beta_t$ or $\mu \beta_t$.
The parameter $0 < \mu < 1$ represents the decreased transmission rate from unreported infected individuals, who are often subclinical or even asymptomatic.
The transmission rate $\beta_t \ge 0$ may change over  time $t$ due to behavioral changes of both susceptible and infected individuals.
Exposed individuals, after an average incubation period of $Z$ days, become reported infected with probability $\alpha_t$ or unreported infected with probability $(1-\alpha_t)$.
The reporting rate $0 < \alpha_t < 1$ may also change over time due to changes in human behavior.
Infected individuals remain infectious for an average period of $D$ days, after which they either recover, or becomes ill enough to be quarantined.
They therefore no longer infect other individuals, and the model does not track their frequency.
The model is described by the following equations:

\begin{equation} \label{eq:model}
\begin{aligned}
\frac{dS}{dt} & = -\beta_t S \frac{I_r}{N} - \mu \beta_t S \frac{I_u}{N} \\
\frac{dE}{dt} & = \beta_t S \frac{I_r}{N} + \mu \beta_t S \frac{I_u}{N}  - \frac{E}{Z} \\
\frac{dI_r}{dt} & = \alpha_t \frac{E}{Z} - \frac{I_r}{D} \\
\frac{dI_u}{dt} & = (1-\alpha_t) \frac{E}{Z} - \frac{I_r}{D} .
\end{aligned}
\end{equation}
The initial numbers of exposed $E(0)$ and unreported infected $I_u(0)$ are considered model parameters, whereas the initial number of reported infected is assumed to be zero $I_r(0)=0$, and the number of susceptible is $S(0)=N-E(0)-I_u(0)$.
This model is inspired by \citet{Li2020} and \citet{Pei2020}, who used a similar model with multiple regions and constant transmission $\beta$ and reporting rate $\alpha$ to infer \covid dynamics in China and the continental US, respectively.

%%% Likelihood %%%
\paragraph*{Likelihood function.}
% TODO update
For a given vector $\theta$ of model parameters the \emph{expected} cumulative number of reported infected individuals ($I_r$) until day $t$ is, following~\autoref{eq:model},
\begin{equation} \label{eq:Yt}
Y_t(\theta) = \int_{0}^{t}{\alpha_s \frac{E(s)}{Z} \; ds}, \quad Y_0 = 0.
\end{equation}
We assume that reported infected individuals are confirmed and therefore observed in the daily case report of day $t$ with probability $p_t$ (note that an individual can only be observed once, and that $p_t$ may change over time, but $t$ is a specific date rather than the time elapsed since the individual was infected).
We denote by $X_t$ the number of confirmed cases in day $t$, and by $\tilde{X}_t$ the cumulative number of confirmed cases until day $t$,
\begin{equation} \label{eq:Xsumt}
\tilde{X}_t=\sum_{i=1}^{t}X_i.
\end{equation}
Therefore, at day $t$ the number of reported infected yet-to-be confirmed individuals is
$(Y_t(\theta) - \tilde{X}_{t-1})$.
We therefore assume that $X_t$ conditioned on $\tilde{X}_{t-1}$ is Poisson distributed,
\begin{equation} \label{eq:Xt} \begin{aligned}
\Big(X_1 \mid \theta \Big) & \sim \mathit{Poi}\big( Y_1(\theta) \cdot p_1 \big), \\
\Big(X_t \mid \tilde{X}_{t-1}, \theta \Big) & \sim 
\mathit{Poi}\Big( \big(Y_t(\theta) - \tilde{X}_{t-1}\big) \cdot p_t \Big), \quad t>1.
\end{aligned}\end{equation}

@@

%%% Model selection %%%
\paragraph*{Model comparison.}
We perform model selection using two methods.
First, we compute WAIC (widely applicable information criterion)~\citep{gelman2013bayesian},
\begin{equation} \label{eq:WAIC}
\begin{aligned}
\mathit{WAIC}(\theta, \vec{X}) &= -2\log\mathbb{E}[\mathcal{L}(\theta \mid \vec{X})] + 2\mathbb{V}[\log\mathcal{L}(\theta \mid \vec{X})]
\end{aligned}
\end{equation}
where $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance operators taken over the posterior distribution $P(\theta \mid \vec{X})$.
We compare models by reporting their relative WAIC; lower is better~(\autoref{table:WAIC}).
A minority of MCMC chains that fail to fully converge can lead to overestimation of the variance (the second term in \autoref{eq:WAIC}).
Therefore, we exclude from the WAIC computation chains with mean log-likelihood that is three standard deviations or more from the overall mean.

We also apply posterior predictive plots, in which we sample 1,000 parameter vectors from the posterior distributions $P(\theta \mid \vec{X})$ and use them to simulate the SEIR model (\autoref{eq:model}).
We plot these simulated dynamics together with the date~(\autoref{fig:ppc}).
Both the accuracy (i.e. overlap of data and prediction) and the precision (i.e. the compactness of the predictions) are good ways to visually compare models.

%%% Source code %%%
\paragraph*{Source code.} 
We use Python 3 with the NumPy, Matplotlib, SciPy, Pandas, Seaborn, and emcee packages.
All source code will be publicly available under a permissive open-source license at \href{http://github.com/yoavram-lab/EffectiveNPI}{github.com/yoavram-lab/EffectiveNPI}.
Files containing samples from the posterior distributions will be deposited on FigShare.

%%% Fig - tau summary %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.6\textwidth]{../figures/Fig-tau-summary.pdf}
    \caption{
    \textbf{Official vs. effective start of non-pharmaceutical interventions.}
    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPI is shown for different regions. The effective NPI dates in Italy and Wuhan are significantly delayed compared to the official dates, whereas in Denmark, France, Spain, and Germany, the effective date is earlier than the official date.
	$\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}. $\tau^*$ is the last NPI date, see \autoref{table:NPI_dates}. Thin and bold lines show 95\% and 75\% credible intervals, respectively (area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
    }
    \label{fig:tau-summary}
\end{figure}

@@

@@

We compared the posterior predictive plots of a model with a free $\tau$ with those of a model with $\tau$ fixed at $\tau^*$ and without $\tau$. The model with free $\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}). 
When we compared the models using WAIC~(\autoref{table:WAIC}), the model with a free parameter for the start of the NPI was better than the other models in 8 out of 12 of the regions; the exceptions are Austria, Belgium, Norway, and United Kingdom.

@@

% Table - WAIC %%%
\begin{table}[h]
\centering
\pgfplotstabletypeset[
    col sep=comma,
    string type,
    every head row/.style={before row=\hline,after row=\hline},
    every last row/.style={after row=\hline},
    ]{../figures/Table-WAIC.csv}
\caption{
\textbf{WAIC values for the different models.}
	WAIC (widely applicable information criterion)~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}. WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy.
}
\label{table:WAIC}
\end{table}

@@

Similarly, in Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be several days layer at Feb~2 ($\pm$2.85 days 95\% CI \autoref{fig:late}).

@@

%%% Early
\paragraph*{Early effective start of NPIs.}
In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (2:1) with a crowd of 50,965 in Seville.
A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ on Mar~8 or~9 ($\pm$1.08 95\%CI), rather than Mar~14 (\autoref{fig:early}).

@@

%%% Assessment
\paragraph*{Effect of late and early effect of NPIs on real-time assessment.}

@@

However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a serial interval of 4-7 days and neglect the late or early population response to the NPI.
This is illustrated in \autoref{fig:fig4} using data and parameters from Italy.
Here, a lockdown is officially ordered on Mar~10 ($\tau^*$), but its late effect on the transmission dynamics starts on Mar~16 ($\hat{\tau}$). If health officials assume the dynamics to immediately change at $\tau^*$, they will expect the number of cases be within the red lines (posterior predictions assuming $\tau=\tau^*$).
This leads to a significant underestimation, which might be interpreted by officials as ineffectiveness of NPIs, leading to further escalations.
However, the number of cases will actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which corresponds well to the real data.

%%% Fig - Prediction %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.5\textwidth]{../figures/Fig-prediction.pdf}
    \caption{
    \textbf{Late effective start of NPIs leads to under-estimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data, line: time moving average). 
    Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution). 
    Shaded box illustrates a serial interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model under-estimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are estimates for Italy (\autoref{table:estimated-params}).
    } 
    \label{fig:fig-prediction}
\end{figure}

@@

For example, in Italy and Wuhan, China, the effective start of the lockdowns seems to have occurred more than five days after the official date (\autoref{fig:late}).
This difference might be explained by low compliance: In Italy, for example, the government intention to lockdown Northern provinces leaked to the public, resulting in people leaving those provinces~\citep{Gatto2020}.
Late effect of NPIs might also be due to the time required by both the government and the citizens to organize for a lockdown, and for the new guidelines to diffuse in the population.

In contrast, in most investigated countries (e.g., Spain and France), we infer reduced transmission rates even before official lockdowns were implemented (\autoref{fig:early}).
An early effective date might be due to adoption of social distancing and similar behavioral adaptations in parts of the population, maybe in response to increased risk perception due to domestic or international \covid-related reports.
This finding may also suggest that severe NPIs, such as lockdowns, were unnecessary, and that less extreme measures adopted by the population could have been sufficient for epidemic control.
These less extreme measures may have been implemented due to government recommendations, media coverage, and social networks, rather than official NPIs.
Indeed, the evidence supports a change in transmission dynamics (i.e. a model with free $\tau$) even for Sweden~(\autoref{fig:ppc}), in which a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}, suggesting that lockdowns may not be necessary if other NPIs are adopted early enough during the outbreak~\citep{Banholzer2020}.

Attempts to asses the effect of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of \covid~\citep{Gatto2020}.
However, the late and early effects we have estimated can confuse these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:fig4}).

We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 9 out of 12 regions (\autoref{table:WAIC}).
It will be interesting to check if the evidence favors a model with \emph{two} change-points, rather than one. 
Two such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), an NPI followed by a relaxation of the intervention, or a mix of NPIs and other events, such as weather, or domestic and international events that affect risk perception.

As several countries (e.g. Austria, Israel) begin to relieve lockdowns and ease restrictions, we expect similar delays and advances to occur: in some countries people will begin to behave as if restrictions were eased even before the official date, and in some countries people will continue to self-restrict even after restrictions are officially removed.

@@

\section*{References}
\nolinenumbers
\bibliographystyle{Science}
%\bibliography{/Users/yoavram/Documents/library}
\bibliography{ms}

@@

>>>>

% NatBib
\usepackage[super,comma]{natbib}
%\usepackage[numbers,super,sort&compress]{natbib}
%\usepackage[round,colon,authoryear]{natbib}
\renewcommand{\bibsection}{}
%\renewcommand{\bibfont}{\small}

@@

During February and March 2020, several countries implemented non-pharmaceutical interventions, such as school closures and lockdowns, with variable schedules to control the \covid pandemic caused by the \sars virus.
Overall, these interventions seem to have successfully reduced the spread of the pandemic.
We hypothesise that the official and effective start date of such interventions can significantly differ, for example due to slow adoption by the population, or due to unpreparedness of the authorities and the public.
We fit an SEIR model to case data from~12 countries to infer the effective start dates of interventions and contrast them with the official dates.
We find both late and early effects of interventions. For example, Italy implemented a nationwide lockdown on Mar~11, but we infer the effective date on Mar~16 ($\pm$0.47 days 95\% CI). In contrast, Spain announced a lockdown on Mar~14, but we infer an effective start date on Mar~8 or~9 ($\pm$ 1.08 days 95\% CI).
We discuss potential causes and consequences of our results.

@@

The \covid pandemic has resulted in implementation of extreme non-pharmaceutical interventions (NPIs) in many affected countries. These interventions, from social distancing to lockdowns, are applied in a rapid and widespread fashion.
NPIs are designed and assessed using epidemiological models, which follow the dynamics of the viral infection to forecast the effect of different mitigation and suppression strategies on the levels of infection, hospitalization, and fatality.
These epidemiological models usually assume that the effect of NPIs on disease transmission begins at the officially declared date~\citep{Li2020,Gatto2020,Flaxman2020}.

Adoption of public health recommendations is often critical for effective response to infectious diseases, and has been studied in the context of HIV~\citep{Kaufman2014} and vaccination~\citep{Dunn2015,Wiyeh2018}, for example.
However, behavioural and social change does not occur immediately, but rather requires time to diffuse in the population through media, social networks, and social interactions. 
Moreover, compliance to NPIs may differ between different interventions and between people with different backgrounds.
For example, in a survey of 2,108 adults in the UK during Mar 2020, \citet{Atchison2020} found that those over 70 years old were more likely to adopt social distancing than young adults (18-34 years old), and that those with lower income were less likely to be able to work from home and to self-isolate.
Similarly, compliance to NPIs may be impacted by personal experiences. \citet{Smith2020} have surveyed 6,149 UK adults in late April and found that people who believe they have already had \covid are more likely to think they are immune, and less likely to comply with social distancing measures. 
Compliance may also depend on risk perception as perceived by the the number of domestic cases or even by reported cases in other regions and countries.
Interestingly, the perceived risk of \covid infection has likely caused a reduction in the number of influenza-like illness cases in the US starting from mid-February~\citep{Zipfel2020}.

Here, we hypothesize that there is a significant difference between the official start of NPIs and their adoption by the public and therefore their effect on transmission dynamics.
We use a \textit{Susceptible-Exposed-Infected-Recovered} (SEIR) epidemiological model and \textit{Markov Chain Monte Carlo} (MCMC) parameter estimation framework to infer the effective start date of NPIs from publicly available \covid case data in several geographical regions.
We compare these estimates to the official dates, and find that they include both late and early effects of NPIs on \covid transmission dynamics.
We conclude by demonstrating how differences between the official and effective start of NPIs can confuse assessments of the effectiveness of the NPIs in a simple epidemic control framework.

@@

%%% Data %%%%
\paragraph*{Data.} 
We use daily confirmed case data $\vec{X}=(X_1, \ldots, X_T)$ from 12~regions. These incidence data summarise the number of individuals $X_t$ tested positive for \sars RNA (using RT-qPCR) at each day $t$.
Data for Wuhan, China retrieved from \citet{Pei2020}, data for 11 European countries retrieved from \citet{Flaxman2020}. 
Regions in which there were multiple sequences of days with zero confirmed cases (e.g. France), we cropped the data to begin with the last sequence so that our analysis focuses on the first sustained outbreak rather than isolated imported cases. 
For dates of official NPI dates see \autoref{table:NPI_dates}.

@@

Susceptible ($S$) individuals become exposed due to contact with reported or unreported infected individuals ($I_r$ or $I_u$) at a rate $\beta_t$ or $\mu \beta_t$.
The parameter $0 < \mu < 1$ represents the decreased transmission rate from unreported infected individuals, who are often subclinical or even asymptomatic~\citep{Thompson2020,Ferretti2020}.
The transmission rate $\beta_t \ge 0$ may change over  time $t$ due to behavioural changes of both susceptible and infected individuals.
Exposed individuals, after an average incubation period of $Z$ days, become reported infected with probability $\alpha_t$ or unreported infected with probability $(1-\alpha_t)$.
The reporting rate $0 < \alpha_t < 1$ may also change over time due to changes in human behavior.
Infected individuals remain infectious for an average period of $D$ days, after which they either recover, or becomes ill enough to be quarantined.
They therefore no longer infect other individuals, and the model does not track their frequency.
The model is described by the following equations:

\begin{equation} \label{eq:model}
\begin{aligned}
\frac{dS}{dt} & = -\beta_t S \frac{I_r}{N} - \mu \beta_t S \frac{I_u}{N} \\
\frac{dE}{dt} & = \beta_t S \frac{I_r}{N} + \mu \beta_t S \frac{I_u}{N}  - \frac{E}{Z} \\
\frac{dI_r}{dt} & = \alpha_t \frac{E}{Z} - \frac{I_r}{D} \\
\frac{dI_u}{dt} & = (1-\alpha_t) \frac{E}{Z} - \frac{I_r}{D} \\
\frac{dR}{dt} & = \frac{I_r}{D} + \frac{I_r}{D} .
\end{aligned}
\end{equation}
The initial numbers of exposed $E(0)$ and unreported infected $I_u(0)$ are free model parameters (i.e. estimated during model fitting), whereas the initial number of reported infected is assumed to be zero $I_r(0)=0$, and the number of susceptible is $S(0)=N-E(0)-I_u(0)$.
This model is inspired by \citet{Li2020} and \citet{Pei2020}, who used a similar model with multiple regions and constant transmission $\beta$ and reporting rate $\alpha$ to infer \covid dynamics in China and the continental US, respectively.

%%% Likelihood %%%
\paragraph*{Likelihood function.}
% TODO update
For a given vector $\theta$ of model parameters the \emph{expected} cumulative number of reported infected individuals ($I_r$) until day $t$ is, following~\autoref{eq:model},
\begin{equation} \label{eq:Yt}
Y_t(\theta) = \int_{0}^{t}{\alpha_s \frac{E(s)}{Z} \; ds}, \quad Y_0 = 0.
\end{equation}
We assume that reported infected individuals are confirmed and therefore observed in the daily case report of day $t$ with probability $p_t$ (note that an individual can only be observed once, and that $p_t$ may change over time, but $t$ is a specific date rather than the time elapsed since the individual was infected).
We denote by $X_t$ the \emph{observed} number of confirmed cases in day $t$, and by $\tilde{X}_t$ the cumulative number of confirmed cases until end of day $t$,
\begin{equation} \label{eq:Xsumt}
\tilde{X}_t=\sum_{i=1}^{t}X_i.
\end{equation}
Therefore, at day $t$ the number of reported infected yet-to-be confirmed individuals is
$(Y_t(\theta) - \tilde{X}_{t-1})$.
We therefore assume that $X_t$ conditioned on $\tilde{X}_{t-1}$ is Poisson distributed,
\begin{equation} \label{eq:Xt} \begin{aligned}
\Big(X_1 \mid \theta \Big) & \sim \mathit{Poi}\big( Y_1(\theta) \cdot p_1 \big), \\
\Big(X_t \mid \tilde{X}_{t-1}, \theta \Big) & \sim 
\mathit{Poi}\Big( \big(Y_t(\theta) - \tilde{X}_{t-1}\big) \cdot p_t \Big), \quad t>1.
\end{aligned}\end{equation}

@@

%%% Model selection %%%
\paragraph*{Model comparison.}
We perform model selection using two methods.
First, we compute WAIC (widely applicable information criterion)~\citep{gelman2013bayesian},
\begin{equation} \label{eq:WAIC}
\begin{aligned}
\mathit{WAIC}(\theta, \vec{X}) &= -2\log\mathbb{E}[\mathcal{L}(\theta \mid \vec{X})] + 2\mathbb{V}[\log\mathcal{L}(\theta \mid \vec{X})]
\end{aligned}
\end{equation}
where $\mathbb{E}[\cdot]$ and $\mathbb{V}[\cdot]$ are the expectation and variance operators taken over the posterior distribution $P(\theta \mid \vec{X})$.
We compare models by reporting their relative WAIC; lower is better~(\autoref{table:WAIC}).
A minority (<5\%) of MCMC chains that fail to fully converge can lead to overestimation of the variance (the second term in \autoref{eq:WAIC}).
Therefore, we exclude from the WAIC computation chains with mean log-likelihood that is three standard deviations or more from the overall mean. %TODO

We also plot posterior predictions: we sample 1,000 parameter vectors from the posterior distribution $P(\theta \mid \vec{X})$, use these parameter vectors to simulate the SEIR model (\autoref{eq:model}), and plot the simulated dynamics~(\autoref{fig:ppc}).
Both the accuracy (i.e. overlap of data and prediction) and the precision (i.e. the compactness of the predictions) are good ways to visually compare models.

%%% Source code %%%
\paragraph*{Source code.} 
We use Python 3 with the NumPy, Matplotlib, SciPy, Pandas, Seaborn, and emcee packages.
All source code will be publicly available under a permissive open-source license at \href{http://github.com/yoavram-lab/EffectiveNPI}{github.com/yoavram-lab/EffectiveNPI}.
Samples from the posterior distributions will be deposited on FigShare.

@@

%%% Fig - tau summary %%%
\begin{figure}[b!]
    \centering
	\includegraphics[width=0.6\textwidth]{../figures/Fig-tau-summary.pdf}
    \caption{
    \textbf{Official vs. effective start of non-pharmaceutical interventions.}
    	The difference between $\hat{\tau}$ the effective and $\tau^*$ the official start of NPI is shown for different regions. The effective NPI dates in Italy and Wuhan are significantly delayed compared to the official dates, whereas in Denmark, France, Spain, and Germany, the effective date is earlier than the official date.
	$\hat{\tau}$ is the posterior median, see \autoref{table:estimated-params}. $\tau^*$ is the last NPI date, see \autoref{table:NPI_dates}. Thin and bold lines show 95\% and 75\% credible intervals, respectively (i.e. area in which $P(|\tau - \hat{\tau}| \mid \vec{X}) = 0.95$ and $0.75$.)
    }
    \label{fig:tau-summary}
\end{figure}

@@

We compared the posterior predictive plots of a model with a free $\tau$ with those of a model with $\tau$ fixed at $\tau^*$ and without $\tau$. The model with free $\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}). 
When we compared the models using WAIC~(\autoref{table:WAIC}), the model with a free parameter for the start of the NPI was preferred over the other models in 8 out of 12 of the regions (although only narrowly for 5 of the 8); the exceptions are Austria, Belgium, Norway, and United Kingdom.

@@

@@

Similarly, in Wuhan, China, a lockdown was ordered on Jan 23~\citep{Li2020}, but we estimate the effective start of NPIs to be more than a week later at Feb~2 ($\pm$2.85 days 95\% CI \autoref{fig:late}).

@@

%%% Early
\paragraph*{Early effective start of NPIs.}
In contrast, in some regions we estimate an effective start of NPIs $\hat{\tau}$ that is \emph{earlier} then the official date $\tau^*$ (\autoref{fig:tau-summary}).
In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a  football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (2:1) with a crowd of 50,965 in Seville.
A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
Nevertheless, we estimate the effective start of NPI $\hat{\tau}$ on Mar~8 or~9 ($\pm$1.08 days 95\%CI), rather than Mar~14 (\autoref{fig:early}).

@@

%%% Assessment
\paragraph*{Consequences of late and early effect of NPIs on real-time assessment.}

@@

However, a significant difference between the beginning of the intervention and the effective change in transmission rates can invalidate assessments that assume a serial interval of 4-7 days and neglect the late or early population response to the NPI.
This is illustrated in \autoref{fig:prediction} using data and parameters from Italy.
Here, a lockdown is officially ordered on Mar~10 ($\tau^*$), but its late effect on the transmission dynamics starts on Mar~16 ($\hat{\tau}$). If health officials assume the dynamics to immediately change at $\tau^*$, they will expect the number of cases be within the red lines (posterior predictions assuming $\tau=\tau^*$).
This leads to a significant underestimation, which might be interpreted by officials as ineffectiveness of NPIs, leading to further escalations.
However, the number of cases will actually follow the blue lines (posterior predictions using $\tau=\hat{\tau}$), which corresponds well to the real data.

%%% Fig - Prediction %%%
\begin{figure}[h]
    \centering
	\includegraphics[width=0.5\textwidth]{../figures/Fig-prediction.pdf}
    \caption{
    \textbf{Late effective start of NPIs leads to under-estimation of daily confirmed cases.}
    Real number of daily cases in Italy in black (markers: data, line: time moving average). 
    Model posterior predictions are shown as coloured lines (1,000 draws from the posterior distribution). 
    Shaded box illustrates a serial interval of seven days.
    \textbf{(A)} Using the official date $\tau^*$ for the start of the NPI,  the model under-estimates the number of cases seven days after the start of the NPI.
    \textbf{(B)} Using the effective date $\hat{\tau}$ for the start of the NPI,  the model correctly estimates the number of cases seven days after the start of the NPI.
    Here, model parameters are estimates for Italy (\autoref{table:estimated-params}).
    } 
    \label{fig:prediction}
\end{figure}

@@

For example, in Italy and Wuhan, China, the effective start of the lockdowns seems to have occurred more than five days after the official date (\autoref{fig:late}).
This difference might be explained by low compliance: In Italy, for example, the government intention to lockdown Northern provinces leaked to the public, resulting in people leaving those provinces~\citep{Gatto2020}.
Late effect of NPIs might also be due to the time required by both the government and the citizens to organise for a lockdown, and for the new guidelines to diffuse in the population.

In contrast, in most investigated countries (e.g., Spain and France), we infer reduced transmission rates even before official lockdowns were implemented (\autoref{fig:early}).
An early effective date might be due to early adoption of social distancing and similar behavioural adaptations in parts of the population. These behavioural adaptations may have a stronger influence than previously thought. In addition, diffusion of these behaviours may occur via media and social networks, rather than official government recommendations and instructions, and may have been influenced by increased risk perception due to domestic or international COVID\=/19-related reports.
Indeed, the evidence supports a change in transmission dynamics (i.e. a model with free $\tau$) even for Sweden~(\autoref{fig:ppc}), in which a lockdown was not implemented\footnote{Sweden banned public events on Mar~12, encouraged social distancing on Mar~16, and closed schools on Mar~18~\citep{Flaxman2020}.}.

Attempts to asses the effect of NPIs~\citep{Flaxman2020,Banholzer2020} generally assume a seven-day delay between the implementation of the intervention and the observable change in dynamics, due to the characteristic serial interval of \covid~\citep{Gatto2020}.
However, the late and early effects we have estimated may bias these assessments and lead to wrong conclusions about the effects of NPIs (\autoref{fig:prediction}).

We have found that the evidence supports a model in which the parameters change at a specific time point $\tau$ over a model without such a change-point in 9 out of 12 regions (\autoref{table:WAIC}).
It could be interesting to check if the evidence favours a model with \emph{two} change-points, rather than one. 
Two such change-points could reflect escalating NPIs (e.g. school closures followed by lockdowns), or an intervention followed by a relaxation.
However, interpretation of such models will be harder, as two change-points can also reflect a mix of NPIs and other events, such as weather, or domestic and international events that affect risk perception.

As several countries (e.g. Austria) begin to relieve lockdowns and ease restrictions, we expect similar delays and advances to occur: in some countries people will begin to behave as if restrictions were eased even before the official date, and in some countries people will continue to self-restrict even after restrictions are officially removed.

@@

\section*{References}
\nolinenumbers
\bibliographystyle{agsm}
\bibliography{/Users/yoavram/Documents/library}
%\bibliography{ms}

@@

% Table - WAIC %%%
\begin{table}[h]
\centering
\pgfplotstabletypeset[
    col sep=comma,
    string type,
    every head row/.style={before row=\hline,after row=\hline},
    every last row/.style={after row=\hline},
    ]{../figures/Table-WAIC_tmp.csv}
\caption{
\textbf{WAIC values for the different models.}
	WAIC (widely applicable information criterion)~\citep{gelman2013bayesian} values for models  with: no $\tau$ at all, \emph{No}; $\tau$ fixed at the official last NPI date $\tau^*$, \emph{Fixed}; and free parameter $\tau$, \emph{Free}. WAIC values are scaled as a deviance measure: lower values imply higher predictive accuracy. Bold emphasise cases in which the \emph{Free} model had the lowest WAIC. $*$ and $**$ mark if the difference was smaller or greater than 2, which is a popular significance level for model comparison~\citep{Kass1995}.
}
\label{table:WAIC}
\end{table}
